### 2018-11-04 Sabanci_talk/Sabanci_talk.tex
More slides

\begin{frame}{Binary Neutron Star Inspirals}{Newtonian evolution}
 Define \bl{chirp mass} \hspace{1.cm} $\red{M_c} \equiv \mu^{3/5} M^{2/5} = \tfrac{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}}$ \\
 %and symmetric mass ratio\qquad $\nu \equiv \tfrac{\mu}{M} =\tfrac{m_1 m_2}{M^2}$\\
 \vspace{1mm}
 which gives, using Kepler's \bl{third law}: $\Omega^2 = G M r^{-3}=\omega^2/4$   
 \begin{footnotesize}
  $$ \boxed{\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}}$$
 \end{footnotesize}\\
%
{\bf Key idea:} \red{energy balance} between $\dot{E}$ and \\
$\tfrac{d}{dt}$ of gravitational binding energy
$ E_b= -\tfrac{G m_1 m_2}{2r} = -\left(\tfrac{G^2 M_c^5 \omega^2}{32}\right)^{1/3} $
%
\[ 
 \dot{E}(f) = -\dot{E}_b(f) \Longrightarrow \boxed{\red{\dot{f}} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, \red{f^{11/3}} }
 \]
\[
 \red{ \dot{f} \propto f^{11/3}\quad(1)}
\]

@@

\begin{frame}{Binary Neutron Star Inspirals}{\todo{Add $r(t)$ plot? }}
Integrate $\dot{f} \sim f^{11/3} \quad \Longrightarrow \quad \int dt \sim \int f^{-11/3}df \quad\Rightarrow\quad t \sim \red{f^{-8/3}}$
\\
\vspace{1mm}
Fix integration constant $\tau = t_\text{coal} - t >0$. \\
\vspace{2mm}
Introduce \red{inspiral time}, i.e, time to BNS \bl{coal}escence
%
\begin{align}
 \red{\tau_\text{insp}(f)} %&= \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,{f^{-8/3}}\nn\\
 &\simeq \bl{16.72\,\text{minutes}} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\red{\left(\f{10\,\text{Hz}}{f}\right)^{8/3}\quad(2)}\nn
\end{align}
Number of GW cycles to coalescence: $\int \tfrac{f}{\dot{f}}df\sim \red{f^{-5/3}}$ %$\int df {f}/\dot{f}$
\begin{align}
 \red{\mathcal{N}_\text{cyc}(f)} %&=\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right]\nn \\
 &\approx 1.605\times \bl{10^4}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\red{ \left(\f{10\,\text{Hz}}{f} \right)^{5/3}}\quad \red{(3)}\nn
\end{align}
%
\begin{tikzpicture}[overlay,remember picture]
\node[draw, red,thick,minimum width=8.2cm,minimum height=1.3cm] (b) at (7.,3.35){};
\node[draw, red,thick,minimum width=10cm,minimum height=1.3cm] (b) at (6.2,1){};
\end{tikzpicture}
%
\[ 
\tfrac{\dot{r}}{r}\sim \tfrac{\dot{f}}{f} \Longrightarrow \red{r(\tau)}=r_i \left(\f{\red{\tau}}{\tau_i}\right)^{\red{1/4}}
\]

@@

\begin{frame}{Gravitational waves from BNS}
TT gauge, multipole, slow-velocity expansions give \todo{add suppl. slide}
\[
 h_{ij}^\text{TT} = \tfrac{1}{D}\tfrac{2G}{c^4} \ddot{Q}_{il}^\text{TT}(t-z/c)
\]
%
 \begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\begin{align}
h_c(t) &=\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \nn\\
\Phi(t) &=\int_{t_i}^t dt' \omega(t') + \text{PN corrections}
\end{align}

Frequecny increases as $f^{11/3}$\\
Amplitude increases as $f^{2/3} \Longrightarrow$ \red{Chirp!}

@@

 \begin{frame}{GW Detectors: Interferometers}
  Geodesic deviation
 \end{frame}

@@

\begin{frame}{Supplement 2: The expression for $h_c(t)$}
Maggiore chapters 1, 3 and 4
\begin{itemize}
 \item Eq. (3.3) expanded to (3.9), (3.14) to (3.34) to (3.55) to (3.59)
 \item Eq. (3.72): $h(t)$ as angular distribution with mass moments $\ddot{M}_{ij}$'s
 \item At that point we get the scaling $h \sim \tfrac{1}{D}\tfrac{G}{c^4}\ddot{Q}_{ij}$.
 \item Eq. (3.275) shows it expanded in tensor harmonics so up to $\partial^2 Y^{\ell m}$
\end{itemize}

@@

>>>>

\begin{frame}{Binary Neutron Star Inspirals}{Newtonian evolution}
 Define \bl{chirp mass} \hspace{1.cm} $\red{M_c} \equiv \mu^{3/5} M^{2/5} = \tfrac{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}}$ \\
 %and symmetric mass ratio\qquad $\nu \equiv \tfrac{\mu}{M} =\tfrac{m_1 m_2}{M^2}$\\
 \vspace{1mm}
 which gives, using Kepler's \bl{third law}: $\Omega^2 = G M r^{-3}=\omega^2/4$   
 \begin{footnotesize}
  $$ \boxed{\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}}$$
 \end{footnotesize}\\
%
{\bf Key idea:} \red{energy balance} between $\dot{E}$ and $\dot{E}_b$\\
 %of gravitational binding energy
\quad where $ E_b= -\tfrac{G m_1 m_2}{2r} = -\left(\tfrac{G^2 M_c^5 \omega^2}{32}\right)^{1/3} $
%
\[ 
 \dot{E}(f) = -\dot{E}_b(f) \Longrightarrow \boxed{\red{\dot{f}} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, \red{f^{11/3}} }
 \]
\[
 \red{ \dot{f} \propto f^{11/3}\quad(1)}
\]

@@

\begin{frame}{Binary Neutron Star Inspirals}{\todo{Add $r(t)$ plot? }}
Integrate $\dot{f} \sim f^{11/3} \quad \Longrightarrow \quad \int dt \sim \int f^{-11/3}df \quad\Rightarrow\quad t \sim \red{f^{-8/3}}$
\\
\vspace{1mm}
Fix integration constant $\tau = t_\text{coal} - t >0$. \\
\vspace{2mm}
Introduce \red{inspiral time}, i.e, time to BNS \bl{coal}escence
%
\begin{align}
 \red{\tau_\text{insp}(f)} %&= \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,{f^{-8/3}}\nn\\
 &\simeq \bl{16.72\,\text{minutes}} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,{\left(\f{10\,\text{Hz}}{\red{f}}\right)^{\red{8/3}}\quad(2)}\nn
\end{align}
Number of GW cycles to coalescence: $\int \tfrac{f}{\dot{f}}df\sim \red{f^{-5/3}}$ %$\int df {f}/\dot{f}$
\begin{align}
 \red{\mathcal{N}_\text{cyc}(f)} %&=\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right]\nn \\
 &\approx 1.605\times \bl{10^4}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}{ \left(\f{10\,\text{Hz}}{\red{f}} \right)^{\red{5/3}}}\quad \red{(3)}\nn
\end{align}
%
\begin{tikzpicture}[overlay,remember picture]
\node[draw, red,thick,minimum width=8.2cm,minimum height=1.3cm] (b) at (7.,3.35){};
\node[draw, red,thick,minimum width=10cm,minimum height=1.3cm] (b) at (6.2,1){};
\end{tikzpicture}
%
\[ 
\tfrac{\dot{r}}{r}\sim \tfrac{\dot{f}}{f} \Longrightarrow \red{r(\tau)}=r_i \left(\f{\red{\tau}}{\tau_i}\right)^{\red{1/4}}
\]

@@

\begin{frame}{Gravitational waves from BNS}
TT gauge, multipole, slow-velocity expansions give \todo{add suppl. slide}
\begin{small}
\[
 h_{ij}^\text{TT} = \tfrac{2}{D}\tfrac{G}{c^4} \ddot{Q}_{ij}^\text{TT}%(t-z/c)
\]
%
 \begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi(t)]\label{eq:hcross_TD},
\end{align}
\end{small}
%
%
where
\vspace{-3mm}
%
\begin{small}
\begin{align}
h_c(t) %&=\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3}\nn\\
&\approx 1.9\times \bl{10^{-23}} \left(\f{40\,\text{Mpc}}{D}\right)\left(\f{M_c}{1.219 M_\odot}\right)^{5/3}\left(\frac{\red{f}}{10\,\text{Hz}}\right)^{\red{2/3}} \nn\\
\Phi(t) &=\int_{t_i}^t dt' \omega(t') + \text{PN corrections}
\end{align}
\end{small}\\
%
Frequency increases as $f^{11/3}$\\
Amplitude increases as $f^{10/3} \Longrightarrow$ \red{Chirp!} \todo{Insert chirp movie}

@@

 \begin{frame}{GW Detectors: \red{Interferometers}}
  Two geodesics separated by $\xi^\mu$:
  %\[
  % \frac{D^2 \xi^\alpha}{D\tau^2} = -R^\alpha_{\beta \gamma\delta}\, \xi^\gamma u^\beta u^\delta 
  %\]
$ \tfrac{d^2 \xi^i}{d\tau^2} = -R^i_{0j0} \xi^j\left(\tfrac{dx^0}{d\tau}\right)^2$ \hfill (detector frame)\\
%\hspace{4cm} $\Longrightarrow \ddot{\xi^i} = -c^2 R^i_{0j0} \xi^j $ \hfill (detector time)\ \\
\vspace{-3mm}
\[
\hspace{4.5cm} \boxed{\ddot{\xi^i} = \frac{1}{2} \ddot{h}_{ij}^\text{TT} \xi^j} \hspace{2.5cm} \text{(detector time)}
\]
%
%
GW along $\mathbf{z}$: 
\[
 h_{ij}^\text{TT} = \left(\begin{array}{cc}
                           h_+ & h_\times \\
                           h_\times & - h_+
                          \end{array}\right) \cos\left(\omega\left( t-\tfrac{z}{c}\right)\right)
\]
Consider mirrors at $(0,0)_1, (L,0)_2$ and $(L,L)_3$
\begin{small}
\begin{align*}
  s_{12} &= \int_1^2 \tfrac{ds}{dx}dx = (x_2-x_1) \left[ 1 + h_+ \cos(\omega t)\right]^{1/2}\hspace{4cm} \\
			      & \approx L \left[ 1+\tfrac{1}{2} h_+ \cos\omega t \right] \\
s_{13} &\approx L \left[ 1-\tfrac{1}{2} h_+ \cos\omega t \right]
\end{align*}
\end{small}
%
\begin{tikzpicture}[overlay,remember picture]
%\node (img0 )[anchor=center,scale=1,opacity=1] at ([shift={(3.8cm,-2.8cm)}]current page.center) {\animategraphics[loop,controls,height=2.39cm]{12}{LIGO_int-}{0}{102}};
\end{tikzpicture}
 \end{frame}

@@

\begin{frame}{Supplement 2: The expression for $h_c(t)$}
Maggiore chapters 1, 3 and 4
\begin{itemize}
 \item Eq. (3.3) expanded to (3.9), (3.14) to (3.34) to (3.55) to (3.59)
 \item Eq. (3.72): $h(t)$ as angular distribution with mass moments $\ddot{M}_{ij}$'s
 \item At that point we get the scaling $h \sim \tfrac{1}{D}\tfrac{G}{c^4}\ddot{Q}_{ij}$.
 \item Eq. (3.275) shows it expanded in tensor harmonics so up to $\partial^2 Y^{\ell m}$
\end{itemize}
\end{frame}

\begin{frame}{Supp: coord. vs. proper distance in TT gauge}
  Geodesic deviation 
  \[
   \frac{D^2 \xi^\alpha}{D\tau^2} = -R^\alpha_{\beta \gamma\delta}\, \xi^\gamma u^\beta u^\delta 
  \]
In TT gauge,\qquad  $\Longrightarrow \quad\left.\tfrac{d^2 \xi^i}{d\tau^2}\right|_{\tau=0} = -\left. \dot{h}_{ij} \tfrac{d\xi^i}{d\tau}\right|_{\tau=0}$\\
i.e., if mirrors are initially at rest w.r.t to each other ( $\left. \tfrac{d\xi^i}{d\tau}\right|_{\tau=0} =0$ )\\
then $\tfrac{d^2 \xi^i}{d\tau^2}=0\quad \forall t$ \\
$\Longrightarrow$ coordinate separation $\xi^i = \text{constant}$ for all time \\
BUT, proper distance is NOT! \\
Consider mirrors at $\mathbf{x}=x_1$ and at $\mathbf{x}=x_2=x_1+\mathbf{\xi}$
\begin{align*}
  s = \int_1^2 \tfrac{ds}{dx}dx &= (x_2-x_1) \left[ 1 + h_+ \cos(\omega\t)\right]^{1/2} \\
			      & \approx L \left[ 1+\tfrac{1}{2} h_+ \cos\omega t \right] 
\end{align*}
 \end{frame}

@@

\begin{frame}{Supplemet 5: Mass-inclination degeneracy}
